---
title: Tim's homework 5 performed with R
author: ''
date: '2020-09-17'
slug: hmwk5
categories:
  - R
  - Teaching
  
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## 1. Obtain the data and read it into a data frame

After skimming the file, I have got two strategies to read the file into a data frame.

1. From the file, the none data parts are labeled with "!", such as below links [1:3], so to get rid of these parts, the remaining are the data requested. 

1. **Idea**: Before and after the data part, two sentences are containing *"matrix_table_begin/end"* (following lines [5:6]). Therefore, I just need to read the information between these two sentences and make it into a data frame.

```{r, comment= " "}
df <- readLines("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE2nnn/GSE2583/suppl/GSE2583_correctly_ordered_Klevecz_GEO_time_series_data.txt")

df[c(1:3,56:59,9393)]

t <- grep('matrix_table',df) # locate the position of the two sentences
read_start <- t[1] + 1 # reservation starts
read_end <- t[2] - 1 # reservation ends

df <- df[read_start:read_end] # keep the data part
df <- as.character(unlist(strsplit(df, "\t"))) # Breaking strings with "\t"   
df <- matrix(df,nrow=33) # Employing all broken strings to form a matrix
df <- data.frame(t(df), stringsAsFactors = F) # Removing factor characteristic of all the strings

# Assigning row/column names to the data frame, and removing the corresponding content
rownames(df) <- df[,33] 
colnames(df) <- df[1,]

df <- df[,-33]
df <- df[-1,]

# Transferring a string data frame to a numeric data frame
df.rownames <- row.names(df)
df <- data.frame(sapply(df, as.numeric))
rownames(df) <- df.rownames
rm(df.rownames)

# Checking the data frame 
head(df[c(1:7)])
```
## 2. Filter the data

remove NAN values
```{r, comment= " "}
dim(df) # row number and column number checking

df <- na.omit(df)

dim(df) 
```

Comparing before and after NAN trimming, basically, there are no NAN values in the data frame

Calculate the variance as well as the coefficient of each row, and then plot the results.

R has its build-in function *var()* to calculate variance, and by applying this function, I build a function to get the coefficient of variance (CV).

```{r}
The_variance <- apply(df, 1, function(x) var(x))
The_coef <- apply(df, 1, function(x) sqrt(var(x))/mean(x))
```

The ploting results showing like figure 1.

```{r,echo=FALSE, fig.cap="The histogram of standard deviations (A) and coefficient of variation (B).", fig.height=4}
par(mfrow = c(1,2))
hist(sqrt(The_variance), breaks = 50,main = 'A. Standard deviations', xlab = NA)
hist(The_coef,breaks = 50, main = 'B. Coefficient of variation', xlab = NA)
```

Is experimental noise independent of absolute magnitude of measurement? **NO**

Accroding to above difination, the coefficient of variation can be calculated with the equation: 

$$c_{v}=\sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(\frac{x_{i}}{\overline{x}}-1)^2}$$

That is to say, the dispersion of obtaining value depends on the ration of measurement value and *mean* value, which illustrates that the noise is relevant to the absolute measurement value. 500 samples have been randomly chosen to show the relationship among *mean, variance* and *CV*. It is obvious to see, with the increase of *mean* and *variance* value, the *CV* value performs a dropping trend. figure 2.  

```{r, echo=FALSE, fig.height=3.5, fig.cap= "The relationship among mean (A), variance (B), and CV (C). With the increases of the former two values, CV has an obvious dropping trend."}
par(mfrow = c(1,3))
df_smpl <- df[sample(1:nrow(df),500),]
df_smpl <- df_smpl[order(rowMeans(df_smpl)),]
df_smpl_var <- apply(df_smpl, 1, function(x) var(x))
df_smpl_coef <- apply(df_smpl, 1, function(x) sqrt(var(x))/mean(x))
plot(rowMeans(df_smpl), xlab = NA, ylab = NA, main = "A. Mean")
plot(sqrt(df_smpl_var), xlab = NA, ylab = NA, main = "B. Variance")
plot(df_smpl_coef, xlab = NA, ylab = NA, main = "C. Coefficience of\n variance")
```

Get the "cleaned" gene subset

Remove all genes except the 1000 with the greatest variation. This will be your "cleaned" data set.

```{r}
# Getting the rank of variantion 
var_dc <- The_variance[order(The_variance,decreasing = T)] 
# Keep the top 1000 ranked genes
df_sub <- df[rownames(df) %in% names(var_dc[1:1000]),] 
```

## 3. Normalize

There is a build-in function *scale()* in R able to perform the normalization that each row has zero mean and unit variance.

```{r}
df_sub_norm <- t(scale(t(df_sub)))
```

## 4. Perform SVD

There is a build-in function *svd()* in R able to perform the SVD. I used the defualt setting to perform the SVD analysis. The reustls contain:

+ d, a vector containing the singular values of matrix, sorted decreasingly;

+ u, a matrix whose columns contain the left singular vectors of matrix;

+ v, a matrix whose columns contain the right singular vectors of matrix.

Following lines are the *head(3)* of each content.

```{r, comment= " "}
SVD <- svd(df_sub_norm,nv = 10)

sapply(SVD, function(x) head(x,3))
```

Plot the dispersion of the singular values (figure 3).

```{r,echo=FALSE,fig.cap="The dispersion of singular values"}
plot(SVD$d, xlab = expression(i), ylab = expression(S[i]), main = "Dispersion of the singular values")
```

Recall that ...

The eigenvalues of the covariance matrix are related to the squares of the singular values (figure 4A & 4B). R has a build-in function *prcomp()* to performs a principal components analysis on the given data matrix and returns the results as an object of class. From the analysis results below, from the forth component we can get 80\% of the variance, see the  the cumulative variance plot (figure 4C).

```{r}
wdbc.pr <- prcomp(df_sub_norm)

cumpro <- cumsum(wdbc.pr$sdev^2 / sum(wdbc.pr$sdev^2))
which(cumpro > 0.8)
```

```{r, echo=FALSE, fig.height=8, fig.cap="The plot of the squares of the singular values (A), the plot eigenvalues of the covariance matix (B), and the cumulative variance plot(C)"}
layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE))
plot(SVD$d^2, main = "A. Plot of the squares of\n the singular values", ylab = 'squares of the singular values', xlab = NA)

# Screeplot of the Eigenvalues of the first 15 PCs (left) & Cumulative variance plot (right)
plot(wdbc.pr, type = "l", npcs = 32, main = "B. Plot of the PCs")


plot(cumpro, xlab = "PC #", ylab = "Amount of explained variance", main = "C.                                Cumulative variance plot                                  ")
abline(v = 4, col="blue", lty=5)
abline(h = 0.8, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC4"),
       col=c("blue"), lty=5, cex=0.6)
```

Project the data onto the first 3 principal components.(figure 5)

```{r,echo=F, comment=FALSE,message=FALSE,fig.cap="Data distribution on the first 3 principal components, {v(1),v(2)} (A), {v(2),v(3)} (B), {v(1),v(3)} (C), as well as the highlited genes."}
pc_for_plot <- as.data.frame(wdbc.pr$x)

library(ggplot2)
library(ggpubr)

p1 <- ggplot(pc_for_plot,aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.2, size = 3) +
  theme_minimal()+ scale_colour_discrete(name  ="Gene")+ theme(legend.position="none")

p2 <- ggplot(pc_for_plot,aes(x = PC3, y = PC2)) +
  geom_point(alpha = 0.2, size = 3) + theme_minimal()+ scale_colour_discrete(name  ="Gene")+ theme(legend.position="none")

p3 <- ggplot(pc_for_plot,aes(x = PC1, y = PC3)) +
  geom_point(alpha = 0.2, size = 3) +
  theme_minimal()+ scale_colour_discrete(name  ="Gene")+ theme(legend.position="none")

ggarrange(p1, p2, p3,
          labels = c("A", "B", "C"),
          ncol = 2, nrow = 2,
          heights = c(6,6),
          common.legend = TRUE, 
          legend="bottom")
```

Pick a gene with high projection onto the first three principal components. Plot its expression across data sets (figure 6).

```{r,echo=FALSE,fig.height=8.5, fig.cap= "The genes with highest projection in the first three principal component"}
par(mfrow=c(3,1))

plot(as.numeric(pc_for_plot[abs(pc_for_plot$PC1) == max(abs(pc_for_plot$PC1)),]),
     xlab = NA, 
     ylab = 'projection', 
     main = paste(row.names(pc_for_plot[abs(pc_for_plot$PC1) == max(abs(pc_for_plot$PC1)),]),"in PC1"),
     xaxt = "n")
abline(v=1,col="red",lty=5,cex=0.3)

plot(as.numeric(pc_for_plot[abs(pc_for_plot$PC2) == max(abs(pc_for_plot$PC2)),]),
     xlab = NA, 
     ylab = 'projection', 
     main = paste(row.names(pc_for_plot[abs(pc_for_plot$PC2) == max(abs(pc_for_plot$PC2)),]),"in PC2"),
     xaxt = "n")
abline(v=2,col="red",lty=5,cex=0.3)

plot(as.numeric(pc_for_plot[abs(pc_for_plot$PC3) == max(abs(pc_for_plot$PC3)),]),
     xlab = "PC#", 
     ylab = 'projection', 
     main = paste(row.names(pc_for_plot[abs(pc_for_plot$PC3) == max(abs(pc_for_plot$PC3)),]),"in PC3"))
axis(1, at = seq(1, 32, by = 1), las=1)
abline(v=3,col="red",lty=5,cex=0.3)
```
